---
title: "Final_Project"
author: "Michael, Leo, and Dylan"
date: "May 5th, 2025"
output: word_document
---

## Date introduction:
Data consists of the date, location, richter, and deaths associated with earthquakes.

The data set is derived from randomly selected samples of notable earthquakes between 1902 - 1999.

A total of 123 observations have been recorded.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(openintro)
library(ggplot2)
openintro::earthquakes
```

## The variables(total 7):
* Year (numerical) - year the earthquake took place
* Month (categorical - ordinal) - month the earthquake took place
* Day (numerical) - day the earthquake took place
* Richter (numerical) - magnitude of earthquake using richter scale
* Area (categorical) - City or geographic location of earthquake
* Region (categorical) - country or countries if the earthquake occurred on border
* Deaths (numerical) - approximate number of deaths caused buy earthquake


## Research questions:
Our primary question of interest: Do Americas and Asia have different average earthquake death tolls?

* Null hypothesis: There is no difference between the average death toll from earthquakes in the Americas vs. Asia.

* Alternative hypothesis: There is a difference between the average death toll from earthquakes in the Americas vs. Asia.

* Type 1 error: We reject H₀ (say continents are different), but in reality H₀ is true. Which means we wrongly claim that some continents have different death rates, and maybe governments overreact or misallocate disaster resources.

* Type 2 error: We fail to reject H₀ (say all continents are similar), but in reality, some are different. We may miss a real difference — some continents really are more dangerous, but no action is taken.

* Type 2 error is worse, if we miss real differences, we might not prepare properly in high-risk areas — which could lead to more unnecessary deaths.

* Supporting questions: Is there another variable impacting the amount of deaths other than region? Can the richter have any reflection on the amount of deaths?


As shown in the graph below, richter does have an effect on the amount of deaths, with the most deaths hovering around mid to high 7 richter. From this data, it is safe to assume that richter plays a role in the average death tolls per continent.
```{r}
#
deaths_per_richter <- earthquakes %>%
  group_by(richter) %>%
  summarize(total_deaths = sum(deaths))

ggplot(deaths_per_richter, aes(x = factor(richter), y = total_deaths)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Total Earthquake Deaths by Richter",
       x = "Richter",
       y = "Total") +
  theme(axis.text.x = element_text(angle = 25))
```

* Expected Outcome: We expect the continents to have a different average number of deaths.


## Summarize Average Deaths by Region and Main Analyses
```{r}
#divide data into continents
continent_data <- c(
  "Guatemala" = "Americas",
  "Russia" = "Asia",
  "Turkey" = "Asia",
  "India" = "Asia",
  "Ecuador" = "Americas",
  "Taiwan" = "Asia",
  "United States" = "Americas",
  "Chile" = "Americas",
  "Asia" = "Asia", 
  "Italy" = "Europe",
  "Iran" = "Asia",
  "China" = "Asia",
  "Japan" = "Asia",
  "Nicaragua" = "Americas",
  "Armenia-Azerbaijan" = "Asia",
  "India-Nepal" = "Asia",
  "Pakistan" = "Asia",
  "Romania" = "Europe",
  "Argentina" = "Americas",
  "Peru" = "Americas",
  "Turkmenistan" = "Asia",
  "Tajikistan" = "Asia",
  "Algeria" = "Africa",
  "Afghanistan" = "Asia",
  "Morocco" = "Africa",
  "Yugoslavia" = "Europe",
  "Chine" = "Asia",
  "New Guinea" = "Oceania",
  "Philippines" = "Asia",
  "Iran-Turkey" = "Asia",
  "Mexico" = "Americas",
  "El Salvador" = "Americas",
  "Colombia-Ecuador" = "Americas",
  "Armenia" = "Asia",
  "Pakistan-Afghanistan" = "Asia",
  "Indonesia" = "Asia",
  "Colombia" = "Americas",
  "Afghanistan-Tajikistan" = "Asia"
)


earthquakes <- earthquakes %>%
  mutate(
    # if table has NA value for deaths, make it 0
    deaths = ifelse(is.na(deaths), 0, deaths),
    
    #make new column in table
    continent = continent_data[region]
    )


avg_deaths <- earthquakes %>%
  group_by(continent) %>%
  summarise(count = n(),
            avg_deaths = mean(deaths),
            .groups = "drop")

# View the summary table
avg_deaths
```


Based on our data and forms, we can try to create a bar chart. According to the bar chart, it will show a visual comparison of the average number of deaths per earthquake in each continent, can also show which continent seems more dangerous on average.

Bar Chart: Visualize Average Deaths
```{r}
# Bar chart of average deaths
ggplot(avg_deaths, aes(x = reorder(continent, -avg_deaths), y = avg_deaths)) +
  geom_bar(stat = "identity", fill = "grey") +
  labs(title = "Average Deaths Per Earthquake by Continent",
       x = "Region", y = "Average Number of Deaths") +
  theme(axis.text.x = element_text(angle = 25))
```
The plot below illustrates the distribution of data contained in the Asian and American continents. The dot plots represent that there is a high level of variation in the data. Specifically, the deaths per earhtquake can greatly vary.

```{r}
#divide data into seperate data sets

#asia data set
data_Asia <- earthquakes %>%
  filter(continent == "Asia") %>%
  select(deaths)

#Americas data set
data_Americas <- earthquakes %>%
  filter(continent == "Americas") %>%
  select(deaths)

plot_data <- bind_rows(
  data.frame(deaths = data_Asia$deaths, continent = "Asia"),
  data.frame(deaths = data_Americas$deaths, continent = "Americas")
)

#Examine data set via dot plot
ggplot(plot_data, aes(x = deaths, y = continent)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.7, color = "black") +
  labs(title = "Dot Plot of Earthquake Deaths by Continent",
       x = "Number of Deaths",
       y = "Continent") 
```


After checking the dot plot, we conclude that there is a noticable amount of variation in the data. While the data is independent, it is likely that it is not normal.
To prove that the data is not normalized, the shapiro test is used. As shown below, the p-values are less than 0.05, which suggests both sets of data are not normal.
```{r}
shapiro.test(data_Americas$deaths)
shapiro.test(data_Asia$deaths)
```

Although not normal, the data is tested through a t.test. While the p-value looks big enough to reject null hypothesis, the data is very noisy (not normalized), making it hard to prove anything. As seen below, the confidence intervals are extremely wide, and the difference in average deaths might come down to randomness.
```{r}
# t.test

t.test(data_Americas$deaths, data_Asia$deaths)
```



## Conclusion:
Based on the data sets that are tested, we conclude that the data does not meet the conditions to be tested in a t.test(). The p-value generated by the t.test() is unreliable (since the data sets are not normalized), and should not be considered for our conclusion. Given this, there is not enough evidence to reject the null hypothesis, so average number of deaths per continent does not differ.
